AlphaEarth Embeddings Deliver France-Wide Deciduous--Evergreen Maps Without GPU Overheads

### 2025-11-08 article/manuscript/article.tex
added writer agent

\title[Towards universal forest mapping]{Towards Universal Forest Mapping: Foundation Embeddings Outperform Hand-Crafted Harmonics for Deciduous–Evergreen Classification}
%je mentionnerai le capteur dans le titre Mapping accuracy for evergreen and deciduous forests using AlphaEarth embeddings vs. sentinel 2 time series. ou Comparing model performances with AlphaEarth embeddings vs. sentinel 2 time series for classifying evergreen and decduous forests pas besoin de mettre ‘in France’

@@

\begin{abstract}
Forest agencies need annual deciduous–evergreen maps to monitor carbon budgets, disturbance recovery, and biodiversity compliance, yet most lack a workflow that scales beyond bespoke campaigns. Existing options demand a compute-or-craft trade-off: either engineer harmonic descriptors with local phenology expertise or train deep networks on GPU fleets.
We evaluate pre-computed AlphaEarth embeddings—the state-of-the-art foundation model now hosted on Google Earth Engine—as a third path that keeps operations lightweight while encoding neighbourhood context that temporal harmonics miss. Using 14.1 million labelled pixels grouped into 639 eco-region-balanced tiles, we train logistic regression, linear support-vector machines, and eco-region-weighted Random Forests on matched 14-dimensional feature sets: physics-informed harmonics (HARM) and AlphaEarth embeddings (EMB).
Across the three learners, EMB lifts accuracy by $2.8 \pm 0.4$ percentage points, macro-F1 by $3.7 \pm 0.5$, halves expected calibration error (0.033 vs 0.059), and reduces national map fragmentation by 68\,\% relative to HARM. A 2023 model transfers to 2018–2022 embeddings with macro IoU between 0.786 and 0.801, showing that the embeddings track inter-annual drift without re-fitting.
By removing bespoke compute and feature engineering, AlphaEarth shifts the bottleneck to label coverage, letting agencies focus on ground truth collection rather than infrastructure.

\keyFont{\section{Keywords:} deciduous-evergreen, AlphaEarth embeddings, harmonic analysis, Random Forest, logistic regression, support vector machine, phenology}
\end{abstract}

@@

% Teaser figure pinned within Introduction
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{images/sat_emb_classes.png}
    \caption{South‑east of France: (left) true‑color satellite view; (middle) learned embedding visualization (AlphaEarth dimensions A46, A18, A05 mapped to RGB)—notice how embeddings sharpen phenological structure without manual feature design; (right) classification probabilities where orange indicates deciduous and blue evergreen. This teaser previews the central finding: embeddings encode spatial context that harmonics miss.}
    \label{fig:teaser_sat_emb_class}
\end{figure}

High-resolution, annually refreshed deciduous–evergreen maps underpin carbon accounting, disturbance response, and biodiversity reporting \citep{Zhu2014,Zhao2019,Verbesselt2010a,Verbesselt2010b,Kennedy2010,Kennedy2018}. Yet most agencies still rely on legacy inventories or continental products that update slowly and blur key forest types. The Copernicus Dominant Leaf Type map separates broadleaf from conifer but merges evergreen oaks with pines \citep{EU2024a}. BD Forêt V2 provides detailed stand labels yet reflects field campaigns from 2007–2018 \citep{IGN2024}. Delivering a current nationwide map therefore remains a bespoke effort rather than an operational service.

Bespoke projects confront a compute-or-craft dilemma. Physics-informed workflows fit harmonic curves to Sentinel-2 indices, extracting amplitudes, phases, and residual variances that summarise seasonal behaviour \citep{Inglada2017,Li2023,Bolton2020}. The descriptors are interpretable and appealing to domain experts, but the tuning effort scales with every eco-region, sensor update, and disturbance regime.

Deep networks promise automatic feature learning \citep{Low2020,Xie2024FoundationEffective}, yet they demand GPU clusters, large labelled datasets, and machine-learning operations that many agencies cannot maintain. The result is a compute-or-craft bottleneck that still blocks operational deployment.

Foundation embeddings offer a third path by shifting the constraint. Pre-trained models distributed through Google Earth Engine remove both the GPU requirement and the fragility of hand-crafted harmonics, concentrating effort on the one ingredient that cannot be outsourced: ground-truth labels. These embeddings encode spatial neighbourhood context alongside temporal phenology, a capability that traditional harmonic descriptors miss. Figure~\ref{fig:teaser_sat_emb_class} previews how embeddings sharpen structure without manual feature design.

Satellite pretraining initiatives such as SatMAE, Prithvi-EO, and AlphaEarth learn multi-sensor, multi-temporal representations that transfer across tasks \citep{Cong2022,Szwarcman2024PrithviEO2,AlphaEarth2025}. AlphaEarth, a state-of-the-art foundation model, is particularly compelling because it publishes annual 64-dimensional embeddings at 10\,m resolution via Google Earth Engine. The vectors have already supported cross-border vegetation typing and disturbance analyses \citep{Houriez2025AEFDataGen,Seydi2025AlphaEarthBurnedArea}, yet no study has benchmarked them against carefully engineered harmonics under matched conditions.

We address that gap with a France-wide benchmark that evaluates harmonics and embeddings on equal footing. Our dataset aggregates 14.1 million labelled forest pixels from four sources, groups them into 639 non-overlapping 2.5\,km tiles, and balances sampling across the eleven GRECO eco-regions. Two compact 14-dimensional feature sets feed three lightweight classifiers—logistic regression, linear support-vector machines, and eco-region-weighted Random Forests. HARM denotes the physics-informed Sentinel-2 harmonic descriptors, and EMB denotes the fold-selected AlphaEarth subset. Tile-grouped cross-validation measures accuracy and calibration, ancillary drivers explain spatial heterogeneity, temporal experiments test whether a frozen 2023 embedding model transfers to 2018–2022 vintages, and national comparisons situate the outputs against Copernicus DLT and BD Forêt.

We hypothesise that pre-computed embeddings outperform engineered harmonics for deciduous–evergreen mapping while remaining operationally lightweight. We test this claim through accuracy and calibration analyses, spatial coherence and environmental driver assessments, and temporal robustness experiments that include comparisons with national inventories. The remainder of the paper details the data and processing pipeline, reports the results, and concludes with operational guidance and remaining limitations.

@@

\subsection{Study area and eco-regional stratification}
\label{sec:greco}
We require spatially representative training data that span the main forest gradients of mainland France and Corsica. Temperate lowlands, montane belts, and Mediterranean shrublands sit within short distances, so we rely on the eleven \emph{Grandes Régions Écologiques} (GRECO) defined by the French National Forest Inventory \citep{IGN2013GRECO}. The GRECO zones distil climate and forest structure transitions from Atlantic mixed forests through semi-continental plateaus to Mediterranean evergreen stands and Alpine massifs.

We partition the national forest mask into 639 non-overlapping 2.5\,km\,$\times$\,2.5\,km tiles (Figure~\ref{fig:training_tiles}). Each tile contains at least ten inventory or plot-labelled pixels and is separated from its neighbours by 5\,km to limit spatial autocorrelation. Once the reference quota is met, remaining forest pixels in each tile are populated with buffered BD Forêt V2 polygons so that both feature families share an identical footprint.

To keep training and evaluation aligned with the true forest extent we apply eco-region weights. For each region \(r\) we compute an effective forest area \(A_r\) by multiplying the GRECO polygons by the mean forest cover ratio reported by the inventory and summing them to obtain \(\sum_j A_j\). With \(N\) labelled pixels overall and \(n_r\) drawn from region \(r\), the per-pixel weight becomes \(w_r = (A_r / \sum_j A_j) / (n_r / N)\). Regions covering 22\,\% of forests but only 21\,\% of pixels, such as the Oceanic Southwest, are scaled by \(w_r \approx 1.06\); over-represented plains drop to \(w_r \approx 0.89\). We normalise the weights so that \(\sum_i w_i = N\), which keeps rare mountain forests from being drowned out by the plains while avoiding over-sampling common stands. The resulting weights range from 0.89 in densely sampled regions to 1.48 in rare mountainous tiles. A small fraction of coastal pixels lies outside the GRECO mask and is excluded from stratified summaries.

The GRECO framework also provides ecological descriptors that contextualise the spatial behaviour reported in Section~\ref{sec:results}. Atlantic bocage and plantation mosaics dominate Regions A and F, semi-continental oak–beech plateaus span Regions B and C, montane belts extend across the Central Massif, Alps, and Pyrenees (Regions G–I), while drought-prone evergreen maquis defines the Mediterranean mainland and Corsica (Regions J and K). Table~\ref{tab:greco_summary} summarises the climate, forest structure, and training sample counts for each region.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{p{0.8cm}p{3.6cm}p{3.9cm}p{4.6cm}p{1.7cm}}
        \toprule
        \textbf{Code} & \textbf{Region (English name)} & \textbf{Climate and terrain} & \textbf{Dominant forest structure} & \textbf{Training pixels} \\
        \midrule
        A & Greater Crystalline and Oceanic West & Humid Atlantic plains and low hills, dense bocage networks & Oak–chestnut coppice with maritime pine and Sitka spruce plantations & 0.49\,M \\
        B & Semi-Oceanic North Centre & Loess plateaus and chalk cuestas of the Paris Basin & Pedunculate/sessile oak with hornbeam or beech; Scots/Corsican pine on sandy soils & 2.39\,M \\
        C & Greater Semi-Continental East & Ardennes and Lorraine uplands with colder winters & Beech–fir and spruce on mesic plateaus, mixed oak lowlands & 2.80\,M \\
        D & Vosges & Steep crystalline range with high precipitation & Silver fir–beech high forests with spruce and Douglas-fir plantations & 0.39\,M \\
        E & Jura & Limestone plateaus under cool montane climate & Calcareous beech–fir mosaics and mixed spruce on karst slopes & 0.20\,M \\
        F & Oceanic Southwest & Landes coastal plain and Atlantic piedmont & Maritime pine estates interleaved with humid oak, alder, and chestnut stands & 2.91\,M \\
        G & Central Massif & Volcanic plateaus and valleys with montane climate & Beech–fir belts, chestnut groves, Scots/Douglas pine on poorer soils & 1.92\,M \\
        H & Alps & Sharp elevational gradients and deep glacial valleys & Deciduous foothills grading to spruce–larch–fir subalpine belts & 0.65\,M \\
        I & Pyrenees & Atlantic–Mediterranean transition, steep valleys & Oak/beech foothills with fir–spruce upper slopes and Mediterranean pine on south-facing flanks & 0.56\,M \\
        J & Mediterranean & Coastal ranges and plateaus with intense summer drought & Evergreen holm and cork oak maquis, Aleppo and maritime pine stands & 1.43\,M \\
        K & Corsica & Crystalline massif with rugged relief & Lowland evergreen maquis and extensive Laricio pine forests above 900\,m & 0.35\,M \\
        \bottomrule
    \end{tabular}
    \caption{Summary of the eleven GRECO eco-regions used for sampling and evaluation. Training pixel counts correspond to the supervised dataset and sum to 14.1 million samples (rounded to two decimals in millions).}
    \label{tab:greco_summary}
\end{table}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{images/tiles_2_5_km_final_visualization.png}
    \caption{Distribution of the 639 eco-region balanced training tiles. Warm colours indicate Mediterranean and montane eco-regions, while cool colours denote Atlantic and semi-continental domains. Each tile contains at least ten labelled pixels, is separated from its neighbours by 5\,km, and serves as an indivisible unit during cross-validation.}
    \label{fig:training_tiles}
\end{figure}

\subsection{Reference labels}
The supervised dataset aggregates 14{,}086{,}937 forest pixels (10\,m resolution) drawn from four complementary sources. (i) PureForest provides 135\,000 lidar-guided patches with expert-verified dominant species, mainly covering monospecific stands in southern France \citep{gaydon2024pureforestlargescaleaeriallidar}. (ii) The RENECOFOR long-term monitoring network supplies plot inventories with tree measurements above 15\,m height collected between 2019 and 2020 \citep{ulrich:hal-03444393}. (iii) The Tree Position Calibration campaign geolocates dominant trees with airborne lidar to refine field GPS positions and species attribution \citep{ONF,IGN_LiDARHD}. (iv) BD Forêt V2 contributes mapped stands ( > 5{,}000\,m$^2$) updated between 2005 and 2019 \citep{IGN2024}. We apply a 100\,m negative buffer to BD Forêt polygons to mitigate edge drift before rasterisation. Each pixel inherits its eco-region and tile identifier; categorical attributes (phenology, genus, species, source, acquisition year) are encoded via consistent look-up tables. Deciduous pixels represent 75.5\,\% of the samples (10{,}639{,}124 pixels), evergreens 24.5\,\% (3{,}447{,}813 pixels). Buffered BD Forêt polygons supply 88.6\,\% of the pixels, while in-situ inventories anchor 11.4\,\%, ensuring we retain authoritative field labels within every eco-region.

\subsection{Feature extraction}
\subsubsection{Sentinel-2 harmonic descriptors}
We reconstruct Sentinel-2 Level-2A surface reflectances from Google Earth Engine monthly MEDIAN composites, applying the s2cloudless cloud-probability mask with a 75\,\% threshold, removing QA60 bits 10 and 11 (opaque and cirrus clouds), and discarding scenes with more than 95\,\% cloudy pixels. No gap filling or scene classification layer is applied. For each tile we derive the vegetation indices NDVI, EVI, NBR, and the SWIR ratio (CRSWIR). The annual signal of each index is fitted with two sinusoidal harmonics through ordinary least squares,
\[
  x(t) = C + \sum_{k=1}^{2} \big[a_k \cos\!\big(2\pi k t/T\big) + b_k \sin\!\big(2\pi k t/T\big)\big], \quad T = 1\,\text{year}.
\]
We recover offsets, amplitudes, phases, and residual variance; phases are transformed into their sine and cosine components to avoid wrap-around discontinuities. These features are purely temporal: each pixel's time series is modelled independently and ignores the surrounding context. Figure~\ref{fig:harmonics_decomposition} illustrates how the two-harmonic model decomposes an NDVI trajectory into interpretable amplitude and phase components while keeping residuals explicit. Recursive feature elimination (Section~\ref{subsubsec:rfe}) retained 14 descriptors that balance interpretability and cross-validated accuracy (Supplementary Section~S1). These descriptors (Table~\ref{tab:harmonic14}) span amplitude, phase, offset, and residual components across four indices: NDVI first-harmonic amplitude captures seasonal greenness pulses; NBR amplitude reflects moisture and structural contrast; CRSWIR offset encodes baseline water content. We denote this 14-dimensional subset as \textbf{HARM}.

>>>>

\title[AlphaEarth embeddings for deciduous-evergreen mapping]{AlphaEarth Embeddings Deliver France-Wide Deciduous--Evergreen Maps Without GPU Overheads}

@@

\begin{abstract}
Remote sensing workflows face a compute versus craft dilemma: either engineer features with domain expertise or train deep networks on GPU infrastructure. Foundation embeddings promise a third path, but controlled benchmarks against carefully tuned baselines are lacking.
We test whether precomputed AlphaEarth embeddings—state-of-the-art foundation models on Google Earth Engine—deliver superior features for phenological classification when classifiers and training data are held constant. Using deciduous–evergreen mapping in France as our controlled testbed, we train logistic regression (LR), linear support-vector machines (SVM), and Random Forests (RF) on 14.1 million pixels with two matched 14-dimensional feature sets: physics-informed harmonics (HARM) and AlphaEarth embeddings (EMB).
Embeddings improve accuracy by $2.8 \pm 0.4$ percentage points and macro-F1 by $3.7 \pm 0.5$ across all classifiers. They halve calibration error (0.033 vs 0.059) and reduce map fragmentation by 68\,\%. A frozen 2023 model transfers to 2018–2022 with macro-IoU between 0.79 and 0.80. Gains concentrate in humid, mixed-stand mosaics where spatial context matters most; Mediterranean evergreen-dominated regions show minimal uplift.
\textbf{The primary contribution of this work is a rigorous feature benchmark demonstrating that foundation embeddings provide superior features under fair experimental conditions.} The nationwide deciduous–evergreen map serves as the application proof: it shows that embedding advantages translate to operational deployment at national scale while also delivering an ecologically meaningful product. The operational bottleneck for agencies therefore shifts away from building complex models and toward their core expertise: curating high-quality ground-truth data.

\keyFont{\section{Keywords:} foundation models, AlphaEarth embeddings, feature engineering, harmonic analysis, Random Forest, phenology, deciduous-evergreen}
\end{abstract}

@@

% Teaser figure pinned within Introduction
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{images/sat_emb_classes.png}
    \caption{Southeast France showing three views: (left) true-color satellite image; (middle) AlphaEarth embedding visualization (dimensions 46, 18, 5 as RGB) revealing phenological patterns without manual design; (right) classification map where orange shows deciduous forest and blue shows evergreen. Key observation: embeddings sharpen forest boundaries and separate adjacent stands that appear similar in temporal signals alone. This preview illustrates the core insight—embeddings encode spatial neighbourhood context that pixel-by-pixel harmonics miss.}
    \label{fig:teaser_sat_emb_class}
\end{figure}

% Claim: Foundation embeddings provide superior features; deciduous mapping demonstrates this
% Evidence: \citep{AlphaEarth2025,Cong2022,Szwarcman2024PrithviEO2}
% Warrant: Controlled comparison isolates feature quality from model complexity
% ABT Agreement: Time-series analysis is the accepted standard for vegetation mapping
Foundation models promise to eliminate the feature-engineering bottleneck in Earth observation \citep{AlphaEarth2025,Cong2022,Szwarcman2024PrithviEO2}. \textbf{And} for over a decade, the remote sensing community has agreed that carefully designed time-series analysis—harmonic fits, phenological metrics, or spline smoothing—delivers robust vegetation maps when paired with lightweight classifiers \citep{Zhu2014,Zhao2019,Verbesselt2010a,Verbesselt2010b,Kennedy2010,Kennedy2018}. These approaches are interpretable, computationally accessible, and operationally proven. \textbf{But} they suffer a critical limitation: each pixel is modelled independently, ignoring neighbourhood context. In spatially heterogeneous landscapes where deciduous and evergreen stands intermingle at 10--100\,m scales, temporal features alone struggle to resolve boundaries. \textbf{Therefore}, pretrained embeddings—which encode both temporal dynamics and spatial context—offer a fundamentally different feature representation. \textbf{This paper answers a single question: do foundation embeddings deliver superior features when evaluated under fair experimental conditions?} We use deciduous-versus-evergreen forest mapping in France as our controlled testbed because the binary phenology distinction is ecologically unambiguous, immediately interpretable by stakeholders, and operationally relevant for carbon accounting, disturbance monitoring, and biodiversity reporting. \textbf{The goal is to prove that embeddings are superior features; the deciduous–evergreen map is the application that demonstrates this claim while also delivering an operationally useful product.} Our scientific contribution is a rigorous feature benchmark proving that AlphaEarth embeddings outperform hand-crafted harmonics when classifiers, training data, and dimensionality are held constant.

% Claim: Embeddings offer third path by encoding spatial context
% Evidence: AlphaEarth accessible via GEE, trained on multi-sensor inputs
% Warrant: Spatial awareness separates embedding features from pixel-wise temporal models
Bespoke remote sensing projects confront a compute versus craft dilemma. Physics-informed workflows fit harmonic curves to Sentinel-2 indices, extracting amplitudes, phases, and residual variances that summarise seasonal behaviour \citep{Inglada2017,Li2023,Bolton2020}. The descriptors are interpretable and appealing to domain experts, but the tuning effort scales with every ecoregion, sensor update, and disturbance regime. Deep networks promise automatic feature learning \citep{Low2020,Xie2024FoundationEffective}, yet they demand GPU clusters, large labelled datasets, and machine learning operations that many organisations cannot maintain.

Foundation embeddings offer a third path. Pretrained models distributed through Google Earth Engine eliminate both local GPU requirements and fragile feature engineering. Effort concentrates on the one ingredient that cannot be outsourced: ground-truth labels. \textbf{The key technical advantage is spatial awareness.} Embeddings encode neighbourhood context alongside temporal patterns, allowing each pixel to incorporate information from its surroundings during inference. Harmonic descriptors model each pixel independently and cannot access this spatial information. Figure~\ref{fig:teaser_sat_emb_class} previews this advantage: embeddings sharpen boundaries between adjacent stands that appear similar in temporal curves alone.

% Claim: AlphaEarth provides state-of-the-art embeddings but lacks controlled benchmarking
% Evidence: \citep{AlphaEarth2025,Houriez2025AEFDataGen,Seydi2025AlphaEarthBurnedArea}
% Warrant: No prior study has isolated embedding quality under matched experimental design
AlphaEarth is a state-of-the-art foundation model for Earth observation \citep{AlphaEarth2025}. It learns from multiple sensors (Sentinel-1/2, Landsat, GEDI, climate data) and publishes annual 64-dimensional embeddings at 10\,m resolution through Google Earth Engine. Similar models (SatMAE, Prithvi-EO) demonstrate transfer learning across tasks \citep{Cong2022,Szwarcman2024PrithviEO2}. AlphaEarth has supported vegetation mapping and burned-area detection \citep{Houriez2025AEFDataGen,Seydi2025AlphaEarthBurnedArea}. However, \textbf{no study has benchmarked these embeddings against carefully tuned harmonics under controlled conditions—matched dimensionality, identical training data, and consistent classifiers.} Without such a comparison, claims about embedding superiority remain anecdotal.

% Claim: Deciduous mapping demonstrates why better features matter operationally
% Evidence: Copernicus DLT limitations, BD Forêt update lag
% Warrant: Current products blur ecologically distinct classes, motivating feature improvement
Deciduous-versus-evergreen phenology is an ideal testbed because existing operational products expose gaps that better features can address. Many agencies still rely on legacy inventories or continental products that update slowly. The Copernicus Dominant Leaf Type (DLT) map separates broadleaf from conifer but does not distinguish deciduous from evergreen broadleaf species; consequently, evergreen oaks in Mediterranean regions are grouped with conifers under the ``needle-leaved'' class \citep{EU2024a}. Visual comparison confirms this limitation: in Provence and coastal Mediterranean zones, sclerophyllous evergreen species such as \textit{Quercus ilex} (holm oak) systematically appear as ``needle-leaved'' in DLT despite being broadleaved trees, creating ecological misclassification that complicates biodiversity and carbon reporting. BD Forêt V2 provides detailed stand labels frozen to 2007–2018 field campaigns \citep{IGN2024}. Delivering current nationwide deciduous–evergreen maps therefore remains a bespoke exercise rather than a routine service. \textbf{If embeddings provide superior features, they should enable routine annual updates with minimal engineering overhead.}

% Claim: National benchmark with Mediterranean as critical test case
% Evidence: GRECO stratification, 14.1M pixels across eleven ecoregions
% Warrant: Mediterranean boundary conditions stress-test the embedding advantage
% CCC Complication: Representativeness is hard when labels are spatially biased
We benchmark harmonics against embeddings across France using 14.1 million labelled pixels from four sources, grouped into 639 non-overlapping 2.5\,km tiles balanced across eleven ecoregions (Table~\ref{tab:greco_summary}). \textbf{Yet simply collecting labels nationwide is insufficient.} Reference datasets suffer spatial autocorrelation: field campaigns concentrate in accessible lowlands, lidar acquisitions favour commercial forests, and photo-interpreted polygons lag behind disturbances. \textbf{Therefore}, we stratify sampling with the GRECO ecoregional framework \citep{IGN2013GRECO} to ensure that Atlantic mosaics, Mediterranean shrublands, and montane belts contribute proportionally (Section~\ref{sec:greco}).

\textbf{This design deliberately includes Mediterranean evergreen-dominated zones as a critical stress test.} In these regions—characterized by intense summer drought, homogeneous holm and cork oak maquis, and low seasonal contrast—spatial context may matter less than in heterogeneous temperate mosaics. \textbf{If embeddings derive their advantage primarily from encoding neighbourhood information, we expect large gains in Atlantic mixed stands and montane gradients where deciduous and evergreen intermingle at 10--100\,m scales, but minimal uplift in Mediterranean shrublands where phenological signals are uniformly weak and evergreen canopies dominate the landscape.} Foreshadowing this outcome now sets a clear boundary condition for when embeddings help and when they do not.

We extract two 14-dimensional feature sets via recursive feature elimination: HARM contains physics-informed Sentinel-2 harmonic descriptors (amplitudes, phases, offsets); EMB contains selected AlphaEarth embedding dimensions. We train three lightweight classifiers on identical folds: logistic regression (LR), linear support-vector machines (SVM), and Random Forests (RF). Lightweight models isolate feature quality from model complexity.

Our evaluation unfolds in two stages. First, we ensure a fair comparison between feature families: recursive feature elimination with cross-validation (RFECV) trims the harmonic descriptors to 14 components; we then apply the same procedure to the 64 embedding channels and retain the best 14 so both models operate under equal dimensionality. We train LR, SVM, and RF on each feature set to measure the overall gain that embeddings provide when everything else is held constant. Second, we focus on the top-performing classifier (RF) to probe spatial coherence, regional drivers, temporal transfer, and alignment with existing national products. This structure lets readers trace the story from balanced features, through global accuracy improvements, to operational viability.

We measure accuracy, calibration, and spatial coherence through tile-grouped cross-validation. We correlate performance with environmental drivers (climate, soil, composition). We test temporal stability by applying a frozen 2023 model to 2018–2022 embeddings. We compare outputs against Copernicus DLT and BD Forêt V2. \textbf{Demonstrating these contrasts on a national deciduous–evergreen application shows how pretrained embeddings can democratise country-scale Earth observation: the primary challenge shifts from complex infrastructure and feature engineering to curating high-quality ground-truth data.}

We hypothesise that precomputed embeddings deliver superior features while remaining operationally lightweight. The remainder of the paper presents methods, evidence, and operational implications.

@@

% CCC Complication: Building a nationally representative benchmark is challenging
% Evidence: Spatial bias in label sources, confidence-coverage tradeoff
% Warrant: Stratified design and data fusion are necessary to isolate feature quality
\textbf{Our goal is to benchmark embedding features against harmonic features under controlled conditions.} Yet building a nationally representative dataset for such a comparison is challenging: reference datasets are spatially biased (field campaigns concentrate in accessible lowlands, lidar acquisitions favour commercial forests, photo-interpreted polygons lag disturbances), and label sources involve a fundamental confidence-versus-coverage tradeoff. High-confidence in-situ inventories are sparse; extensive mapped products carry lower taxonomic certainty and temporal lag. Simply pooling available datasets would over-represent lowland agricultural edges and under-sample Mediterranean shrublands or montane gradients. \textbf{Therefore}, we deliberately stratify sampling with the GRECO ecoregional framework and fuse four complementary label sources to ensure every major ecological gradient contributes proportionally. This design lets us isolate feature quality from sampling artifacts.

Section~\ref{sec:greco} details the ecoregional stratification and label fusion strategy. The following subsections introduce the two feature sets (harmonics and embeddings), the recursive feature elimination that aligns their dimensionality, and the classifiers and metrics used to quantify gains and practical usability.

\subsection{Study area and ecoregional stratification}
\label{sec:greco}
We need folds that cover the main ecological gradients of mainland France and Corsica so that any feature advantage holds nationally, not just in a handful of regions. Temperate lowlands, montane belts, and Mediterranean shrublands sit within short distances, so we rely on the eleven \emph{Grandes Régions Écologiques} (GRECO) defined by the French National Forest Inventory \citep{IGN2013GRECO}. These ecoregions distil the dominant climate and forest structure transitions—from Atlantic mixed forests through semi-continental plateaus to Mediterranean evergreen stands and Alpine massifs—and provide the scaffold for stratified sampling.

We partition the national forest mask into 639 non-overlapping 2.5\,km\,$\times$\,2.5\,km tiles (Figure~\ref{fig:training_tiles}). Each tile must contain at least ten high-confidence reference pixels drawn from in-situ inventories or lidar-guided datasets; this ensures every fold is anchored by authoritative labels. Once that quota is reached we complete the tile with buffered BD Forêt V2 polygons so that both feature families share the same footprint for training. Tiles can touch but never overlap, which keeps cross-validation units distinct. The four reference sources—PureForest lidar stands, the RENECOFOR monitoring plots, the Tree Position Calibration campaign, and BD Forêt V2 polygons—are detailed in the next subsection.

To produce unbiased national summaries we weight tiles to reflect each ecoregion’s forest area. For region \(r\) we compute an effective forest area \(A_r\) by multiplying the GRECO polygons by the mean forest cover ratio reported by the inventory and summing them to obtain \(\sum_j A_j\). With \(N\) labelled pixels overall and \(n_r\) drawn from region \(r\), the per-pixel weight becomes \(w_r = (A_r / \sum_j A_j) / (n_r / N)\). Regions that are under-represented in the sample (e.g., the Oceanic Southwest with 22\,\% of forest but 21\,\% of pixels) receive \(w_r > 1\); over-represented plains are down-weighted. We normalise the weights so that \(\sum_i w_i = N\), preventing dominant lowland regions from drowning out mountainous or Mediterranean tiles. The GRECO descriptors also provide ecological context for the regional analyses reported later (Table~\ref{tab:greco_summary}).

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{p{0.8cm}p{3.6cm}p{3.9cm}p{4.6cm}p{1.7cm}}
        \toprule
        \textbf{Code} & \textbf{Region (English name)} & \textbf{Climate and terrain} & \textbf{Dominant forest structure} & \textbf{Training pixels} \\
        \midrule
        A & Greater Crystalline and Oceanic West & Humid Atlantic plains and low hills, dense bocage networks & Oak–chestnut coppice with maritime pine and Sitka spruce plantations & 0.49\,M \\
        B & Semi-Oceanic North Centre & Loess plateaus and chalk cuestas of the Paris Basin & Pedunculate/sessile oak with hornbeam or beech; Scots/Corsican pine on sandy soils & 2.39\,M \\
        C & Greater Semi-Continental East & Ardennes and Lorraine uplands with colder winters & Beech–fir and spruce on mesic plateaus, mixed oak lowlands & 2.80\,M \\
        D & Vosges & Steep crystalline range with high precipitation & Silver fir–beech high forests with spruce and Douglas-fir plantations & 0.39\,M \\
        E & Jura & Limestone plateaus under cool montane climate & Calcareous beech–fir mosaics and mixed spruce on karst slopes & 0.20\,M \\
        F & Oceanic Southwest & Landes coastal plain and Atlantic piedmont & Maritime pine estates interleaved with humid oak, alder, and chestnut stands & 2.91\,M \\
        G & Central Massif & Volcanic plateaus and valleys with montane climate & Beech–fir belts, chestnut groves, Scots/Douglas pine on poorer soils & 1.92\,M \\
        H & Alps & Sharp elevational gradients and deep glacial valleys & Deciduous foothills grading to spruce–larch–fir subalpine belts & 0.65\,M \\
        I & Pyrenees & Atlantic–Mediterranean transition, steep valleys & Oak/beech foothills with fir–spruce upper slopes and Mediterranean pine on south-facing flanks & 0.56\,M \\
        J & Mediterranean & Coastal ranges and plateaus with intense summer drought & Evergreen holm and cork oak maquis, Aleppo and maritime pine stands & 1.43\,M \\
        K & Corsica & Crystalline massif with rugged relief & Lowland evergreen maquis and extensive Laricio pine forests above 900\,m & 0.35\,M \\
        \bottomrule
    \end{tabular}
    \caption{Summary of the eleven GRECO ecoregions used for sampling and evaluation. Training pixel counts correspond to the supervised dataset and sum to 14.1 million samples (rounded to two decimals in millions).}
    \label{tab:greco_summary}
\end{table}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{images/tiles_2_5_km_final_visualization.png}
    \caption{Distribution of the 639 ecoregion-balanced training tiles. Warm colours indicate Mediterranean and montane ecoregions, while cool colours denote Atlantic and semi-continental domains. Each tile contains at least ten labelled pixels anchored by in-situ references and serves as an indivisible unit during cross-validation.}
    \label{fig:training_tiles}
\end{figure}

\subsection{Reference labels}
% CCC Complication: Nationally representative labels are hard to assemble
% Evidence: Four complementary sources
% Warrant: Fusion strategy balances confidence and coverage
Building a nationally representative and unbiased training dataset is difficult. \textbf{The fundamental complication is that label availability is spatially autocorrelated and skewed:} field inventories deliver verified species identifications and high confidence \textit{but} cover sparse locations concentrated in accessible forests; photo-interpreted products span entire landscapes and deliver extensive training data \textit{but} carry lower taxonomic confidence and may lag temporal updates. Simply pooling available datasets would over-represent lowland agricultural edges and under-sample Mediterranean shrublands or montane gradients. \textit{Therefore}, we deliberately fuse four label sources—balancing confidence against coverage—to satisfy the ecoregional stratification requirements. High-confidence in-situ data (PureForest lidar stands, RENECOFOR plots, Tree Position Calibration) anchor each tile with authoritative ground truth (Section~\ref{sec:greco} mandates at least ten such pixels per tile), while broader mapped products (BD Forêt V2) fill spatial gaps and ensure every ecoregion contributes sufficient training pixels without regional bias. This fusion strategy lets us test both feature families under realistic operational constraints where perfect labels are unavailable but operational decisions still demand national coverage.

The supervised dataset aggregates 14{,}086{,}937 forest pixels (10\,m resolution) from the following sources. (i) PureForest provides 135\,000 lidar-guided patches with expert-verified dominant species, mainly covering monospecific stands in southern France \citep{gaydon2024pureforestlargescaleaeriallidar}. (ii) The RENECOFOR long-term monitoring network supplies plot inventories with tree measurements above 15\,m height collected between 2019 and 2020 \citep{ulrich:hal-03444393}. (iii) The Tree Position Calibration campaign geolocates dominant trees with airborne lidar to refine field GPS positions and species attribution \citep{ONF,IGN_LiDARHD}. (iv) BD Forêt V2 contributes mapped stands (>5{,}000\,m$^2$) updated between 2005 and 2019 \citep{IGN2024}. A 100\,m negative buffer mitigates edge drift before rasterisation. Each pixel inherits its ecoregion and tile identifier; categorical attributes (phenology, genus, species, source, acquisition year) are encoded via consistent look-up tables. Deciduous pixels represent 75.5\,\% of the samples (10{,}639{,}124 pixels), evergreens 24.5\,\% (3{,}447{,}813 pixels). Buffered BD Forêt polygons supply 88.6\,\% of pixels, while in-situ inventories anchor 11.4\,\%, giving every ecoregion a mix of authoritative and widespread labels.

\subsection{Feature extraction}

We compare two complementary families of descriptors. Harmonic features encode seasonal cycles through hand-crafted sinusoidal fits, whereas AlphaEarth embeddings provide pretrained vectors that already mix temporal history with spatial neighbourhood context. The following subsections detail how each family is constructed before they are trimmed to matching dimensionality.
\subsubsection{Sentinel-2 harmonic descriptors}
We start from monthly median composites of Sentinel-2 Level-2A surface reflectances on Google Earth Engine. The s2cloudless probability mask (75\,\% threshold) and QA60 bits remove clouds; scenes with more than 95\,\% cloudy pixels are discarded. This delivers one clean snapshot per month without additional gap filling. We derive four spectral indices that respond to canopy properties: NDVI and EVI (chlorophyll and greenness), NBR (woody biomass and moisture), and the SWIR ratio CRSWIR (water content and lignin).

Seasonal behaviour is summarised with a simple harmonic model. Deciduous forests in France complete one growth-and-senescence cycle per year, typically leaf-off for four to five months. We therefore fit two sine/cosine pairs per index: the first captures the annual component, the second captures a half-year component that models asymmetric spring green-up and autumn decline. Ordinary least squares estimates the coefficients, and phases are expressed through sine and cosine terms to avoid angle discontinuities. From each fit we retain offsets (mean level), amplitudes (seasonal intensity), phases (timing), and residual variance (departures due to disturbance or mixed pixels). These descriptors are purely temporal—each pixel is modelled independently without neighbourhood information.
